---
title: "ST558 Project 1"
author: "Alfonso Straffon"
format: html
editor: visual
---


# Introduction
The goal of this project is to write functions that will *manipulate* and *process* 
data from the [Census Public Use Microdata Sample (PUMS) API](https://www.census.gov/data/developers/data-sets/census-microdata-api.html), notably generic functions 
that will combine, summarize, and plot the queried data. The project will use PUMS 
data from 2021 through 2024.

The PUMS data contains person and household level characteristics for records returned 
by the Census API. Each row represents a sampled record, but the rows do not all carry 
the same population weight. The PWGTP variable indicates how many people in the population 
are represented by each record, which becomes important when summarizing the data.

The variables used in this project include numerical, categorical, time, and geography 
variables. The table below provides a brief description of each variable and the units 
or type of information it contains.

| Variable | Description | Type / Units |
|:---------|:------------|:-------------|
| AGEP | Age | Numerical - years |
| GASP | Monthly gas cost | Numerical - dollars per month |
| GRPIP | Gross rent as a percentage of household income | Numerical - percent |
| JWMNP | Travel time to work | Numerical - minutes |
| PWGTP | Person weight; number of people represented by the record | Numerical - population weight |
| JWAP | Time of arrival at work | Time |
| JWDP | Time of departure for work | Time |
| FER | Gave birth to a child within the past 12 months | Categorical |
| HHL | Household language | Categorical |
| SCH | School enrollment | Categorical |
| SCHL | Educational attainment | Categorical |
| SEX | Sex | Categorical |
| ST / STATE | State code (`ST` in 2021-2022 and `STATE` in 2023-2024) | Geography |
| REGION | Census region | Geography |
| DIVISION | Census division | Geography |
| year | Survey year added when combining multiple years | Categorical in our summaries |


# Initial Setup
Before starting the actual project we first need to: 1) create a github repository, 
2) set up github pages, 3) clone the repo locally as an R project, 4) create a `.qmd` 
file, and 5) push said changes back to github in order to create a webpage. This 
sequence of steps largely followed what we did in Homework 1. Lastly, in the code 
chunk below, we loaded the needed R packages to be used throughout the project. 

Now we are ready to begin!
```{r}
#| message: false
#| warning: false

#needed packages throughout project
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
```

# Part 1: Data Processing

## Parsing Variable Values

### Step 1: Manual setup
First we will manually query the URL and download the data for a single year (2021) 
using the `GET()` function from the `httr` package, then we will parse the data with 
the `content()` function and store it as an R object (list), and lastly will select 
variables of interest. 

```{r}

# Year 2021
# Get 2021 Census variable information
variable_return_2021 <- GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2021 <- content(variable_return_2021)

# Variables needed for 2021
variables_2021 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","ST", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2021 <- parsed_2021$variables[variables_2021]

# Pull the values for each variable
variable_values_2021 <- lapply(variable_info_2021, function(x) x$values)
```

### Step 2: Writing a function to call any year
Next, we write a function that repeats the above process for any given year.
```{r}
# function to get variable values for one year
get_variable_values <- function(year) {
  
  # create URL for selected year
  variable_URL <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )
  
  # get Census variable information
  variable_return <- GET(variable_URL)
  
  # parse JSON into an R object
  parsed_variables <- content(variable_return)
  
  # variables needed depend on year because ST changes to STATE
  if (year %in% c(2021, 2022)) {
    
    variables_needed <- c(
      "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
      "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION"
    )
    
  } else {
    
    variables_needed <- c(
      "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
      "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION"
    )
  }
  
  # keep only the variables needed for the project
  variable_info <- parsed_variables$variables[variables_needed]
  
  # pull the values for each variable
  variable_values <- lapply(
    variable_info,
    function(x) x$values
  )
  
  variable_values
}
```

### Step 3: Call function for years of interest
Now we proceed to run it for 2021, 2022, 2023, and 2024, with results stored in a *.rds* 
file.
```{r}

variable_values_2021 <- get_variable_values(2021)
variable_values_2022 <- get_variable_values(2022)
variable_values_2023 <- get_variable_values(2023)
variable_values_2024 <- get_variable_values(2024)

# Combine yearly variable-value lists into one master list
variable_values <- list(
  "2021" = variable_values_2021,
  "2022" = variable_values_2022,
  "2023" = variable_values_2023,
  "2024" = variable_values_2024
)

# Save list as an .rds file in the data folder
saveRDS(
  variable_values,
  "data/pums_variable_values.rds"
)

# read saved variable information
variable_values <- readRDS("data/pums_variable_values.rds")

```

## Parsing API Data
Now we will write a function that lets a user to query the PUMS API using an API key 
and return the results in a formatted tibble.

###  Step 1: Getting the process to work with base case

```{r}

#key needed
API_key <- "de74f201ac188240a9b31ef280b4980495d05d22"

#base URL
URL_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums"

# Query parameters wrapped in list()
query_2024 <- list(
  get = "AGEP,PWGTP,SEX",
  `for` = "state:37",
  key = API_key
)

# Query the API
census_return_2024 <- GET(
  URL_2024,
  query = query_2024
)

```

### Step 2: Use a helper function to create tibble

The function will do the the following, starting with the raw response:
1) raw response : the API's returned data in machine language
2) rawToChar(): converts machine language to human-readable JSON text
3) fromJSON(): converts the JSON array into an R Object
4) create a tibble

But first we want to run through this process manually to create a tibble before 
automating with a function. This helps illustrate the steps that will be embedded 
in the *helper* function. 

```{r}
# converts returned response to readable JSON text
json_text_2024 <- rawToChar(census_return_2024$content)

# converts the JSON array into an R object
parsed_2024 <- fromJSON(json_text_2024)

# save first row as column names
column_names <- parsed_2024[1, ]

# remove first row and convert remaining data to tibble
census_data_2024 <- as_tibble(
  parsed_2024[-1, ]
)

# assign column names previously saved
names(census_data_2024) <- column_names

#inspect tibble
head(census_data_2024)

```

Now we can proceed with creating the helper function to automate this process.
```{r}
# helper function to convert API response into a tibble
census_tibble <- function(api_return) {
  
  # convert returned response into an R object
  parsed_data <- fromJSON(
    rawToChar(api_return$content)
  )
  
  # save first row as column names
  column_names <- parsed_data[1, ]
  
  # remove first row and convert remaining data to tibble
  census_data <- as_tibble(
    parsed_data[-1, ]
  )
  
  # assign column names previously saved
  names(census_data) <- column_names
  
  # return tibble
  census_data
}

# test helper function
census_data_test <- census_tibble(census_return_2024)

# inspect tibble
head(census_data_test)

```

### Step 3: Write main API Function
Next we *wrap* the helper function into a main function that parses the data, 
but that will also allow users to change the survey year, numerical variables, 
categorical variables, geography level, and geography value, as long as the 
selected inputs are valid.


```{r}
# Default settings for function f(x1,x2,x3,x4,x5) with five inputs
census_query <- function(
  year = 2024,
  numerical_variables = c("AGEP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = NULL
) { # now we add the body of function in between these curly brackets
  
  # check that only one year was entered
  if (length(year) != 1) {
    stop("Only one year can be entered.")
  }
  
  # valid years
  valid_years <- c(2021, 2022, 2023, 2024)
  
  # check that year is valid
  if (!(year %in% valid_years)) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # valid numerical variables
  valid_numerical_variables <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
  )
  
  # check that only valid numerical variables were entered
  if (!all(numerical_variables %in% valid_numerical_variables)) {
    stop("Invalid numerical variable entered.")
  }
  
  # check that PWGTP was included
  if (!("PWGTP" %in% numerical_variables)) {
    stop("PWGTP must be included.")
  }
  
  # check that at least one numerical variable besides PWGTP was included
  if (length(setdiff(numerical_variables, "PWGTP")) < 1) {
    stop("At least one numerical variable other than PWGTP must be included.")
  }
  
  # valid categorical variables
  valid_categorical_variables <- c(
    "FER", "HHL", "SCH", "SCHL", "SEX"
  )
  
  # check that at least one categorical variable was entered
  if (length(categorical_variables) < 1) {
    stop("At least one categorical variable must be included.")
  }
  
  # check that only valid categorical variables were entered
  if (!all(categorical_variables %in% valid_categorical_variables)) {
    stop("Invalid categorical variable entered.")
  }
  
  # valid geography levels
  valid_geographies <- c("Region", "Division", "State")
  
  # check that only one geography level was entered
  if (length(geography) != 1) {
    stop("Only one geography level can be entered.")
  }
  
  # check that geography level is valid
  if (!(geography %in% valid_geographies)) {
    stop("Geography must be Region, Division, or State.")
  }
  
  # determine which geography variable to use
  if (geography == "State") {
    
    # state variable changed names after 2022
    if (year %in% c(2021, 2022)) {
      geography_variable <- "ST"
    } else {
      geography_variable <- "STATE"
    }
    
  } else if (geography == "Region") {
    
    geography_variable <- "REGION"
    
  } else if (geography == "Division") {
    
    geography_variable <- "DIVISION"
  }
  
  # pull allowable geography values from saved variable information
  geography_values <-
    variable_values[[as.character(year)]][[geography_variable]][["item"]]
  
  # Census geography codes are stored as the names
  valid_geography_values <- names(geography_values)
  
  # assign a default geography value if none was entered
  if (is.null(geography_value)) {
    
    if (geography == "State") {
      geography_value <- "37"
    } else if (geography == "Region") {
      geography_value <- "3"
    } else if (geography == "Division") {
      geography_value <- "5"
    }
  }
  
  # convert geography value to character
  geography_value <- as.character(geography_value)
  
  # check that only one geography value was entered
  if (length(geography_value) != 1) {
    stop("Only one geography value can be entered.")
  }
  
  # check that geography value is valid or is "all"
  if (!(geography_value %in% valid_geography_values) &&
      !(tolower(geography_value) == "all")) {
    stop("Invalid geography value entered.")
  }
  
  # build API URL
  API_URL <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")

  # build query parameters
  # combine numerical and categorical variables
  variables_to_get <- c(
    numerical_variables,
    categorical_variables
  )

  # convert variable names to comma-separated format for API
  get_variables <- paste(
    variables_to_get,
    collapse = ","
  )

  # use * if user selected all geography values
  if (tolower(geography_value) == "all") {
    geography_query_value <- "*"
  } else {
    geography_query_value <- geography_value
  }

  # build geography portion of query
  geography_query <- paste0(
    tolower(geography),
    ":",
    geography_query_value
  )

  # build query parameters
  query_parameters <- list(
    get = get_variables,
    `for` = geography_query,
    key = API_key
  )
  
  # run GET()
  api_return <- GET(
    API_URL,
    query = query_parameters
  )
  
  # convert API response to tibble
  census_data <- census_tibble(api_return)
  
  # convert variable types

  # 1) ordinary numerical variables

  # identify time variables
  time_variables <- c("JWAP", "JWDP")

  # numerical variables other than time variables
  regular_numerical_variables <- setdiff(
    numerical_variables,
    time_variables
  )

  # convert regular numerical variables to numeric
  for (variable in regular_numerical_variables) {
    census_data[[variable]] <- as.numeric(census_data[[variable]])
  }

  # 2) time variables

  # identify selected time variables
  selected_time_variables <- intersect(
    numerical_variables,
    time_variables
  )

  # convert selected time variables to midpoint times
  for (variable in selected_time_variables) {
  
    census_data[[variable]] <- hms::as_hms(
      sapply(
        census_data[[variable]],
        time_midpoint,
        year = year,
        variable = variable
      )
    )
  }

  # 3) categorical variables

  for (variable in categorical_variables) {
  
    # pull category values from saved variable information
    category_values <-
      variable_values[[as.character(year)]][[variable]][["item"]]
  
    # convert categorical variable to factor
    census_data[[variable]] <- factor(
      census_data[[variable]],
      levels = names(category_values),
      labels = unlist(category_values)
    )
  }

  # add census class
  class(census_data) <- c("census", class(census_data))

  # return tibble
  census_data
}

census_query()
```


## Combining Years
Next, we write a function that allows the user to query multiple years of Census data. 
This function will use the single-year `census_query()` function created above for each 
selected year and then combine the results into one tibble. We use `lapply()` to take 
each year and apply the single-year function.

```{r}
# function to combine Census data from multiple years...note `years' is plural
census_query_years <- function(
  years,
  numerical_variables = c("AGEP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = NULL
) {
  
  # query Census data for each selected year
  yearly_data <- lapply(
    years,
    function(current_year) {
      
      census_data <- census_query(
        year = current_year,
        numerical_variables = numerical_variables,
        categorical_variables = categorical_variables,
        geography = geography,
        geography_value = geography_value
      )
      
      # add year variable
      census_data$year <- current_year
      
      # return data for current year
      census_data
    }
  )
  
  # combine all years into one tibble
  combined_data <- bind_rows(yearly_data)
  
  # make sure census class is included only once
  class(combined_data) <- unique(c("census", class(combined_data)))
  
  # return combined tibble
  combined_data
}

# test combining two years
census_years_test <- census_query_years(years = c(2023, 2024))

# inspect results
head(census_years_test)
str(census_years_test)
class(census_years_test)

unique(census_years_test$year)

write.csv(
  head(census_years_test, 10000),
  "census_years_test_sample.csv",
  row.names = FALSE

)
```
# Part 2: Writing Generic Functions

## Numeric and Categorical Summaries

Now that we have functions that return Census data for one or multiple years, we
can create a custom `summary()` function for objects with the `"census"` class.
This function will calculate weighted summaries for numerical variables and counts
for categorical variables.
```{r}
# custom summary function for census class
summary.census <- function(
  object,
  numerical_variables = NULL,
  categorical_variables = NULL
) {
  
  # identify time variables to exclude from numerical summaries
  time_variables <- c("JWAP", "JWDP")
  
  # if numerical variables are not specified, use all numeric variables
  if (is.null(numerical_variables)) {
    
    numerical_variables <- names(object)[
      sapply(object, is.numeric)
    ]
    
    # remove variables that should not receive numerical summaries
    numerical_variables <- setdiff(
      numerical_variables,
      c("PWGTP", "year", time_variables)
    )
  }
  
  # if categorical variables are not specified, use factor and character variables
  if (is.null(categorical_variables)) {
    
    categorical_variables <- names(object)[
      sapply(
        object,
        function(x) is.factor(x) || is.character(x)
      )
    ]
    
    # treat year as a categorical variable
    if ("year" %in% names(object)) {
      categorical_variables <- c(categorical_variables, "year")
    }
  }
  
  # create an empty list to store summary results
  summary_results <- list()
  
  # calculate weighted summaries for numerical variables
  for (variable in numerical_variables) {
    
    # pull numerical values and population weights
    numeric_values <- object[[variable]]
    weight_values <- object$PWGTP
    
    # keep rows without missing values
    valid_rows <- !is.na(numeric_values) & !is.na(weight_values)
    
    numeric_values <- numeric_values[valid_rows]
    weight_values <- weight_values[valid_rows]
    
    # calculate weighted mean
    weighted_mean <- sum(
      numeric_values * weight_values
    ) / sum(weight_values)
    
    # calculate weighted standard deviation
    weighted_sd <- sqrt(
      sum(numeric_values^2 * weight_values) / sum(weight_values) -
        weighted_mean^2
    )
    
    # save results for this variable
    summary_results[[variable]] <- c(
      mean = weighted_mean,
      sd = weighted_sd
    )
  }
  
  # calculate weighted counts for categorical variables
  for (variable in categorical_variables) {
    
    # pull categorical values and population weights
    category_values <- object[[variable]]
    weight_values <- object$PWGTP
    
    # keep rows without missing values
    valid_rows <- !is.na(category_values) & !is.na(weight_values)
    
    category_values <- category_values[valid_rows]
    weight_values <- weight_values[valid_rows]
    
    # calculate weighted counts for each category
    weighted_counts <- tapply(
      weight_values,
      category_values,
      sum
    )
    
    # save counts for this variable
    summary_results[[variable]] <- weighted_counts
  }
  
  # return named list of summary results
  summary_results
}

summary(census_years_test)
```
## Graphical Summaries
Next, we create a custom plot() function for Census objects. The function will allow 
the user to specify one categorical variable and one numerical variable, and will 
create a weighted boxplot using `PWGTP`.
```{r}
# custom plot function for census class
plot.census <- function(
  x,
  cat_var,
  num_var,
  ...
) {
  
  # create weighted boxplot
  ggplot(
    x,
    aes(
      x = factor(get(cat_var)),
      y = get(num_var),
      weight = PWGTP
    )
  ) +
    geom_boxplot() +
    labs(
      x = cat_var,
      y = num_var,
      title = paste(num_var, "by", cat_var)
    )
}

plot(
  census_years_test,
  cat_var = "year",
  num_var = "AGEP"
)
```

