---
title: "ST558 Project 1"
author: "Alfonso Straffon"
format: html
editor: visual
---


# Introduction
The goal of this project is to write functions that will *manipulate* and *process* 
data from the Census Public Use Microdata Sample (PUMS) API, notably generic functions 
that will combine, summarize, and plot the queried data. The project will use PUMS 
data from 2021 through 2024.

# Initial Setup
Before starting the actual project we first need to: 1) create a github repository, 
2) set up github pages, 3) cloned the repo locally as an R project, 4) create a `.qmd` 
file, and 5) push said changes back to github in order to create a webpage. This 
sequence of steps largely followed what we did in Homework 1. Lastly, in the code 
chunk below, we loaded the needed R packages to be used throughout the project. 

Now we are ready to begin!
```{r}
#| message: false
#| warning: false

#needed packages throughout project
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
```

# Part 1: Data Processing

## Parsing Variable Values
To download the data we will use the `GET()` function from the `httr` package to 
query the URL for a year of interest, then we will parse the data with the `content()` 
function and store it as an R object (list), and lastly will select variables of 
interest. This process will be repeated for the years 2021, 2022, 2023, and 
2024, with results stored in an *.rds* file.

```{r}

# Year 2021
# Get 2021 Census variable information
variable_return_2021 <- GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2021 <- content(variable_return_2021)

# Variables needed for 2021
variables_2021 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","ST", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2021 <- parsed_2021$variables[variables_2021]

# Pull the values for each variable
variable_values_2021 <- lapply(variable_info_2021, function(x) x$values)

# Year 2022
# Get 2022 Census variable information
variable_return_2022 <- GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2022 <- content(variable_return_2022)

# Variables needed for 2022
variables_2022 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","ST", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2022 <- parsed_2022$variables[variables_2022]

# Pull the values for each variable
variable_values_2022 <- lapply(variable_info_2022, function(x) x$values)

# Year 2023
# Get 2023 Census variable information
variable_return_2023 <- GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2023 <- content(variable_return_2023)

# Variables needed for 2023
variables_2023 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","STATE", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2023 <- parsed_2023$variables[variables_2023]

# Pull the values for each variable
variable_values_2023 <- lapply(variable_info_2023, function(x) x$values)

# Year 2024
# Get 2024 Census variable information
variable_return_2024 <- GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2024 <- content(variable_return_2024)

# Variables needed for 2024
variables_2024 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","STATE", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2024 <- parsed_2024$variables[variables_2024]

# Pull the values for each variable
variable_values_2024 <- lapply(variable_info_2024, function(x) x$values)

# Combine yearly variable-value lists into one master list
variable_values <- list(
  "2021" = variable_values_2021,
  "2022" = variable_values_2022,
  "2023" = variable_values_2023,
  "2024" = variable_values_2024
)

# Save list as an .rds file in the data folder
saveRDS(
  variable_values,
  "data/pums_variable_values.rds"
)


```
## Parsing API Data
Now we will write a function that lets a user to query the PUMS API using an API key 
and return the results in a formatted tibble.

###  Step 1: Getting the process to work with base case

```{r}

#key needed
API_key <- "de74f201ac188240a9b31ef280b4980495d05d22"

#base URL
URL_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums"

# Query parameters wrapped in list()
query_2024 <- list(
  get = "AGEP,PWGTP,SEX",
  `for` = "state:37",
  key = API_key
)

# Query the API
census_return_2024 <- GET(
  URL_2024,
  query = query_2024
)

```

### Step 2: Use a helper function to create tibble

The function will do the the following, starting with the raw response:
1) raw response : the API's returned data in machine language
2) rawToChar(): converts machine language to human-readable JSON text
3) fromJSON(): converts the JSON array into an R Object
4) create a tibble

But first we want to run through this process manually to create a tibble before 
automating with a function.

```{r}
# converts returned response to readable JSON text
json_text_2024 <- rawToChar(census_return_2024$content)

# converts the JSON array into an R object
parsed_2024 <- fromJSON(json_text_2024)

# save first row as column names
column_names <- parsed_2024[1, ]

# remove first row and convert remaining data to tibble
census_data_2024 <- as_tibble(
  parsed_2024[-1, ]
)

# assign column names previously saved
names(census_data_2024) <- column_names

#inspect tibble
head(census_data_2024)

```
Now we can proceed with creating the tibble with a function

```{r}
# Default settings for function f(x1,x2,x3,x4,x5) with five inputs
census_query <- function(
  year = 2024,
  numerical_variables = c("AGEP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = NULL
) { # now we add the body of function in between these curly brackets
  
  # check that only one year was entered
  if (length(year) != 1) {
    stop("Only one year can be entered.")
  }
  
  # valid years
  valid_years <- c(2021, 2022, 2023, 2024)
  
  # check that year is valid
  if (!(year %in% valid_years)) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # valid numerical variables
  valid_numerical_variables <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
  )
  
  # check that only valid numerical variables were entered
  if (!all(numerical_variables %in% valid_numerical_variables)) {
    stop("Invalid numerical variable entered.")
  }
  
  # check that PWGTP was included
  if (!("PWGTP" %in% numerical_variables)) {
    stop("PWGTP must be included.")
  }
  
  # check that at least one numerical variable besides PWGTP was included
  if (length(setdiff(numerical_variables, "PWGTP")) < 1) {
    stop("At least one numerical variable other than PWGTP must be included.")
  }
  
  # valid categorical variables
  valid_categorical_variables <- c(
    "FER", "HHL", "SCH", "SCHL", "SEX"
  )
  
  # check that at least one categorical variable was entered
  if (length(categorical_variables) < 1) {
    stop("At least one categorical variable must be included.")
  }
  
  # check that only valid categorical variables were entered
  if (!all(categorical_variables %in% valid_categorical_variables)) {
    stop("Invalid categorical variable entered.")
  }
  
  # valid geography levels
  valid_geographies <- c("Region", "Division", "State")
  
  # check that only one geography level was entered
  if (length(geography) != 1) {
    stop("Only one geography level can be entered.")
  }
  
  # check that geography level is valid
  if (!(geography %in% valid_geographies)) {
    stop("Geography must be Region, Division, or State.")
  }
  
  # determine which geography variable to use
  if (geography == "State") {
    
    # state variable changed names after 2022
    if (year %in% c(2021, 2022)) {
      geography_variable <- "ST"
    } else {
      geography_variable <- "STATE"
    }
    
  } else if (geography == "Region") {
    
    geography_variable <- "REGION"
    
  } else if (geography == "Division") {
    
    geography_variable <- "DIVISION"
  }
  
  # pull allowable geography values from saved variable information
  geography_values <-
    variable_values[[as.character(year)]][[geography_variable]][["item"]]
  
  # Census geography codes are stored as the names
  valid_geography_values <- names(geography_values)
  
  # assign a default geography value if none was entered
  if (is.null(geography_value)) {
    
    if (geography == "State") {
      geography_value <- "37"
    } else if (geography == "Region") {
      geography_value <- "3"
    } else if (geography == "Division") {
      geography_value <- "5"
    }
  }
  
  # convert geography value to character
  geography_value <- as.character(geography_value)
  
  # check that only one geography value was entered
  if (length(geography_value) != 1) {
    stop("Only one geography value can be entered.")
  }
  
  # check that geography value is valid or is "all"
  if (!(geography_value %in% valid_geography_values) &&
      !(tolower(geography_value) == "all")) {
    stop("Invalid geography value entered.")
  }
  
  # temporary test
  geography_value
}

census_query()
```
