---
title: "ST558 Project 1"
author: "Alfonso Straffon"
format: html
editor: visual
---

# Introduction

The goal of this project is to write functions that will *manipulate* and *process* data from the [Census Public Use Microdata Sample (PUMS) API](https://www.census.gov/data/developers/data-sets/census-microdata-api.html), notably generic functions that will combine, summarize, and plot the queried data. The project will use PUMS data from 2021 through 2024.

The PUMS data contains person and household level characteristics for records returned by the Census API. Each row represents a sampled record, but the rows do not all carry the same population weight. The PWGTP variable indicates how many people in the population are represented by each record, which becomes important when summarizing the data.

The variables used in this project include numerical, categorical, time, and geography variables. The table below provides a brief description of each variable and the units or type of information it contains.

| Variable | Description | Type / Units |
|:-------------------|:------------------------|:--------------------------|
| AGEP | Age | Numerical - years |
| GASP | Monthly gas cost | Numerical - dollars per month |
| GRPIP | Gross rent as a percentage of household income | Numerical - percent |
| JWMNP | Travel time to work | Numerical - minutes |
| PWGTP | Person weight; number of people represented by the record | Numerical - population weight |
| JWAP | Time of arrival at work | Time |
| JWDP | Time of departure for work | Time |
| FER | Gave birth to a child within the past 12 months | Categorical |
| HHL | Household language | Categorical |
| SCH | School enrollment | Categorical |
| SCHL | Educational attainment | Categorical |
| SEX | Sex | Categorical |
| ST / STATE | State code (`ST` in 2021-2022 and `STATE` in 2023-2024) | Geography |
| REGION | Census region | Geography |
| DIVISION | Census division | Geography |
| year | Survey year added when combining multiple years | Categorical in our summaries |

# Initial Setup

Before starting the actual project we first need to: 1) create a github repository, 2) set up github pages, 3) clone the repo locally as an R project, 4) create a `.qmd` file, and 5) push said changes back to github in order to create a webpage. This sequence of steps largely followed what we did in Homework 1. Lastly, in the code chunk below, we loaded the needed R packages to be used throughout the project.

Now we are ready to begin!

```{r}
#| message: false
#| warning: false

#needed packages throughout project
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
```

# Part 1: Data Processing

## Parsing Variable Values

### Step 1: Manual setup

First we will manually query the URL and download the data for a single year (2021) using the `GET()` function from the `httr` package, then we will parse the data with the `content()` function and store it as an R object (list), and lastly will select variables of interest. Doing this manually first helps us understand the process before turning it into a function that can be used for multiple years.

```{r}

# Year 2021
# Get 2021 Census variable information
# GET() from httr sends a request to the API and returns an HTTP response object
variable_return_2021 <- GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# content() from httr parses the response content into an R object
parsed_2021 <- content(variable_return_2021)

# Variables needed for 2021
variables_2021 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","ST", "REGION", "DIVISION")

# use [] to subset the variables list and keep only the variables needed
variable_info_2021 <- parsed_2021$variables[variables_2021]

# Pull the values for each variable
# lapply() is a Base R function that applies the same function to each item in a list
variable_values_2021 <- lapply(variable_info_2021, function(x) x$values)
```

### Step 2: Writing a function to call any year

Next, we write a function that repeats the above process for any given year. This avoids having to manually repeat the same API request and parsing steps for each year.

```{r}
# function to get variable values for one year
get_variable_values <- function(year) {
  # paste0() is a Base R function that joins text together into one character string
  # create URL for selected year
  variable_URL <- paste0("https://api.census.gov/data/",year,"/acs/acs1/pums/variables.json")
  
  # GET() sends the request to the Census API
  variable_return <- GET(variable_URL)
  
  # content() parses the returned response into an R object
  parsed_variables <- content(variable_return)
  
  # if ... else uses a logical condition to decide which code to run
  # %in% checks whether year is 2021 or 2022; if TRUE use ST, otherwise use STATE
  if (year %in% c(2021, 2022)) {
    
    variables_needed <- c(
      "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
      "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION"
    )
    
  } else {
    
    variables_needed <- c(
      "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
      "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION"
    )
  }
  
  # use [] to subset the variables list and keep only the variables needed
  variable_info <- parsed_variables$variables[variables_needed]
  
  # function(x) is an anonymous function; x represents each item in variable_info
  # and x$values pulls the values component from each item
  variable_values <- lapply(variable_info, function(x) x$values)
  
  variable_values
}
```

### Step 3: Call function for years of interest

Now we proceed to run it for 2021, 2022, 2023, and 2024, with results stored in a *.rds* file.

```{r}
# call the function once for each year of interest
variable_values_2021 <- get_variable_values(2021)
variable_values_2022 <- get_variable_values(2022)
variable_values_2023 <- get_variable_values(2023)
variable_values_2024 <- get_variable_values(2024)

# list() is a Base R function that combines objects into a list
# Combine yearly variable-value lists into one master list
variable_values <- list(
  "2021" = variable_values_2021,
  "2022" = variable_values_2022,
  "2023" = variable_values_2023,
  "2024" = variable_values_2024
)

# saveRDS() saves a single R object to an .rds file
# save the master list in the data folder
saveRDS(
  variable_values,
  "data/pums_variable_values.rds"
)

# readRDS() reads the saved R object back into R
# read saved variable information for use later in the project
variable_values <- readRDS("data/pums_variable_values.rds")

```

We now have one saved object containing the variable values for all four years, which we will use later when validating and formatting the Census data returned by the API.

## Parsing API Data

Now we will write a function that lets a user to query the PUMS API using an API key and return the results in a formatted tibble. 

### Step 1: Getting the process to work with base case
We will first run through the query process manually, for a single year, so we can see how the API response is converted into usable R data and then a tibble.

```{r}

#Census API key
API_key <- "de74f201ac188240a9b31ef280b4980495d05d22"

#Base URL for the 2024 PUMS API
URL_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums"

# Query parameters are grouped via list()
query_2024 <- list(
  get = "AGEP,PWGTP,SEX",
  `for` = "state:37", #note`for` is placed in backticks because for is a reserved word in R
  key = API_key)

#GET() sends the request to the Census API using the URL and query parameters
census_return_2024 <- GET(
  URL_2024,
  query = query_2024
)

# converts returned response to readable JSON text
json_text_2024 <- rawToChar(census_return_2024$content)

# converts the JSON array into an R object
parsed_2024 <- fromJSON(json_text_2024)

# save first row as column names
column_names <- parsed_2024[1, ]

# remove first row and convert remaining data to tibble
census_data_2024 <- as_tibble(
  parsed_2024[-1, ]
)

# assign column names previously saved
names(census_data_2024) <- column_names

#inspect tibble
head(census_data_2024)

```

### Step 2: Use a helper function to create tibble

Now we proceed with creating a helper function to automate the process we just completed manually. The process will do the following: 1) the API returns the data as raw content, 2) `rawToChar()` converts the raw content into readable JSON text, 3) `fromJSON()` converts the JSON text into an R object, 4) the first row is saved as the column names, and 5) the remaining data is converted into a tibble.

```{r}
# helper function to convert API response into a tibble
census_tibble <- function(api_return) {
  
  # convert returned response into an R object
  parsed_data <- fromJSON(
    rawToChar(api_return$content)
  )
  
  # save first row as column names
  column_names <- parsed_data[1, ]
  
  # remove first row and convert remaining data to tibble
  census_data <- as_tibble(
    parsed_data[-1, ]
  )
  
  # assign column names previously saved
  names(census_data) <- column_names
  
  # return tibble
  census_data
}


# helper function to convert Census time codes to midpoint times
time_midpoint <- function(code, year, variable) {
  
  # pull the time values for the selected year and variable
  time_values <-
    variable_values[[as.character(year)]][[variable]][["item"]]
  
  # use the Census code to find the corresponding time range
  time_range <- time_values[[as.character(code)]]
  
  # return NA if the time value is not applicable
  if (is.null(time_range) || grepl("^N/A", time_range)) {
    return(hms::as_hms(NA_real_))
  }
  
  # remove periods from a.m. and p.m.
  time_range <- gsub("\\.", "", time_range)
  
  # split the beginning and ending times
  time_parts <- strsplit(time_range, " to ")[[1]]
  
  # convert the beginning and ending times
  start_time <- parse_date_time(
    time_parts[1],
    orders = "I:M p"
  )
  
  end_time <- parse_date_time(
    time_parts[2],
    orders = "I:M p"
  )
  
  # calculate the midpoint of the time range
  midpoint_seconds <- (
    as.numeric(start_time) +
    as.numeric(end_time)
  ) / 2
  
  # return the midpoint in time format
  hms::as_hms(
    midpoint_seconds %% (24 * 60 * 60)
  )
}
```

### Test Section
```{r}
# test helper function
census_data_test <- census_tibble(census_return_2024)

# inspect tibble
head(census_data_test)

# test the time_midpoint() helper using a JWAP code
time_midpoint(
  code = "122",
  year = 2024,
  variable = "JWAP"
)

# test a missing / not-applicable time code... should return NA
time_midpoint(
  code = "0",
  year = 2024,
  variable = "JWAP"
)

```

### Step 3: Write main API Function

Next we *wrap* the helper function into a main function that parses the data, but that will also allow users to change the survey year, numerical variables, categorical variables, geography level, and geography value, as long as the selected inputs are valid. We will first define the default inputs and then add checks to make sure the user has entered allowable values.

The main function will use the following inputs:

- `year`: which survey year to query
- `numerical_variables`: which numerical variables to return
- `categorical_variables`: which categorical variables to return
- `geography`: State, Region, or Division
- `geography_value`: which particular state, region, or division to return

```{r}
# Default settings for function f(x1,x2,x3,x4,x5) with five inputs
census_query <- function(
  year = 2024,
  numerical_variables = c("AGEP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = NULL
) { # now we add the body of function in between these curly brackets
  
  # check that only one year was entered
  # length() returns the number of values in an object
  if (length(year) != 1) {
    # stop() ends the function and returns an error message
    stop("Only one year can be entered.")
  }
  
  # valid years
  valid_years <- c(2021, 2022, 2023, 2024)
  
  # check that year is valid
  if (!(year %in% valid_years)) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # valid numerical variables
  valid_numerical_variables <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
  )
  
  # check that only valid numerical variables were entered
  # all() checks whether every selected variable is valid
  if (!all(numerical_variables %in% valid_numerical_variables)) {
    stop("Invalid numerical variable entered.")
  }
  
  # check that PWGTP was included
  if (!("PWGTP" %in% numerical_variables)) {
    stop("PWGTP must be included.")
  }
  
  # check that at least one numerical variable besides PWGTP was included
  # setdiff() returns values in the first vector that are not in the second
  if (length(setdiff(numerical_variables, "PWGTP")) < 1) {
    stop("At least one numerical variable other than PWGTP must be included.")
  }
  
  # valid categorical variables
  valid_categorical_variables <- c(
    "FER", "HHL", "SCH", "SCHL", "SEX"
  )
  
  #check that at least one categorical variable was entered
  if (length(categorical_variables) < 1) {
    stop("At least one categorical variable must be included.")
  }
  
  #check that only valid categorical variables were entered
  if (!all(categorical_variables %in% valid_categorical_variables)) {
    stop("Invalid categorical variable entered.")
  }
  
  # valid geography levels
  valid_geographies <- c("Region", "Division", "State")
  
  # check that only one geography level was entered
  if (length(geography) != 1) {
    stop("Only one geography level can be entered.")
  }
  
  # check that geography level is valid
  if (!(geography %in% valid_geographies)) {
    stop("Geography must be Region, Division, or State.")
  }
  
  # determine which geography variable to use
  if (geography == "State") {
    
    #state variable changed names after 2022
    if (year %in% c(2021, 2022)) {
      geography_variable <- "ST"
    } else {
      geography_variable <- "STATE"
    }
    
  } else if (geography == "Region") {
    
    geography_variable <- "REGION"
    
  } else if (geography == "Division") {
    
    geography_variable <- "DIVISION"
  }
  
  # pull allowable geography values from saved variable information
  # [[ ]] extracts an individual element from a list
  # as.character() converts year to character to match our named yearly lists
  geography_values <-
    variable_values[[as.character(year)]][[geography_variable]][["item"]]
  
  # Census geography codes are stored as the names
  # names() returns the Census geography codes stored as the names
  valid_geography_values <- names(geography_values)
  
  # assign a default geography value if none was entered
  # is.null() checks whether the user left geography_value unspecified
  if (is.null(geography_value)) {
    
    if (geography == "State") {
      geography_value <- "37"
    } else if (geography == "Region") {
      geography_value <- "3"
    } else if (geography == "Division") {
      geography_value <- "5"
    }
  }
  
  #as.character() converts the geography value to character
  geography_value <- as.character(geography_value)
  
  #check that only one geography value was entered
  if (length(geography_value) != 1) {
    stop("Only one geography value can be entered.")
  }
  
  # check that geography value is valid or is "all"
  # tolower() converts text to lowercase so all, All, and ALL are treated the same
  # && means both logical conditions must be TRUE
  if (!(geography_value %in% valid_geography_values) &&
      !(tolower(geography_value) == "all")) {
    stop("Invalid geography value entered.")
  }
  
  # build API URL using a selected year
  API_URL <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")

  # build query parameters
  # combine numerical and categorical variables into one vector
  variables_to_get <- c(
    numerical_variables,
    categorical_variables
  )

  # convert variable names to comma-separated format for API
  # paste() joins character values together
  # collapse = "," combines the variable names into the comma-separated
  # format required by the Census API
  get_variables <- paste(
    variables_to_get,
    collapse = ","
  )

  # "*" tells the Census API to return all values for the selected geography
  if (tolower(geography_value) == "all") {
    geography_query_value <- "*"
  } else {
    geography_query_value <- geography_value
  }

  # build geography portion of query
  # ":" separates the geography type from its value in the format required by the API
  geography_query <- paste0(
    tolower(geography),
    ":",
    geography_query_value
  )

  # group API query parameters into a list
  query_parameters <- list(
    get = get_variables,
    `for` = geography_query,
    key = API_key
  )
  
  # run GET() to send request to Census API
  api_return <- GET(
    API_URL,
    query = query_parameters
  )
  
  # use our helper function to convert API response to a tibble
  census_data <- census_tibble(api_return)
  
  # convert returned variables to appropriate data types

  # 1) ordinary numerical variables

  # identify time variables
  time_variables <- c("JWAP", "JWDP")

  # setdiff() removes the time variables from the selected numerical variables
  regular_numerical_variables <- setdiff(
    numerical_variables,
    time_variables
  )

  # a for loop repeats the same code for each selected numerical variable
  # as.numeric() converts each variable from character to numeric
  for (variable in regular_numerical_variables) {
    census_data[[variable]] <- as.numeric(census_data[[variable]])
  }

  # 2) time variables

  # intersect() returns values that appear in both vectors
  # identify which time variables were selected by the user
  selected_time_variables <- intersect(
    numerical_variables,
    time_variables
  )

  # convert selected time variables to midpoint times
  for (variable in selected_time_variables) {
    
    # sapply() applies time_midpoint() to each value in the selected variable
    # as_hms() converts the resulting values into a time format
    census_data[[variable]] <- hms::as_hms(
      sapply(
        census_data[[variable]],
        time_midpoint,
        year = year,
        variable = variable
      )
    )
  }

  # 3) categorical variables

  # repeat the conversion for each selected categorical variable
  for (variable in categorical_variables) {
  
    # pull category values from saved variable information
    category_values <-
      variable_values[[as.character(year)]][[variable]][["item"]]
  
    # factor() converts the variable into a categorical factor
    # using the Census codes as levels and descriptions as labels
    census_data[[variable]] <- factor(
      census_data[[variable]],
      levels = names(category_values),
      labels = unlist(category_values)
    )
  }

  # add "census" as the first class while retaining the existing tibble classes
  class(census_data) <- c("census", class(census_data))

   # return completed tibble
  census_data
}
```

### Test Section

```{r}
# test function using all default inputs
census_query()

# query 2023 PUMS data with a broader set of variables
pums_2023 <- census_query(
  year = 2023,
  numerical_variables = c(
    "AGEP", "GASP", "GRPIP", "JWMNP", "PWGTP"
  ),
  categorical_variables = c(
    "FER", "HHL", "SCH", "SCHL", "SEX"
  ),
  geography = "State",
  geography_value = "37"
)

# test main function using a time variable
time_query_test <- census_query(
  year = 2024,
  numerical_variables = c("JWAP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = "37"
)

# inspect returned data and variable types
head(time_query_test)
str(time_query_test)

# keep the first 10,000 rows
pums_sample_10000 <- head(pums_2023, 10000)

# export sample to Excel
writexl::write_xlsx(
  pums_sample_10000,
  "data/pums_sample_10000.xlsx"
)


```
The default function `census_query()` returns 2024 Census data using the default numerical variables `AGEP` and `PWGTP`, categorical variable `SEX`, and the default State geography. The output also shows that the numerical variables have been converted to numeric values and `SEX` has been converted to a factor with the appropriate labels.

In the second test, we query 2023 PUMS data with a broader set of variables. A 10,000-row sample of the returned data is available for inspection here:

[Download the 10,000-row PUMS sample](data/pums_sample_10000.xlsx)

## Combining Years

Next, we write a function that allows the user to query multiple years of Census data. This function will use the single-year `census_query()` function created above for each selected year and then combine the results into one tibble. We use `lapply()` to take each year and apply the single-year function.

```{r}
# function to combine Census data from multiple years
# note that years is plural because multiple years can be entered
census_query_years <- function(
  years,
  numerical_variables = c("AGEP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = NULL
) {
  
  # query Census data for each selected year
  # lapply() applies the same function to each selected year and returns a list
  yearly_data <- lapply(
    years,
    function(current_year) {
      
      # query one year at a time using our census_query() function
      census_data <- census_query(
        year = current_year,
        numerical_variables = numerical_variables,
        categorical_variables = categorical_variables,
        geography = geography,
        geography_value = geography_value
      )
      
      # add a year variable so we know which survey year each row came from
      census_data$year <- current_year
      
      # return data for the current year
      census_data
    }
  )
  
  # combine all years into one tibble
  # bind_rows() from dplyr combines the yearly tibbles by stacking the rows
  combined_data <- bind_rows(yearly_data)
  
  # make sure census class is included only once
  # unique() removes duplicate class names if "census" is already present
  class(combined_data) <- unique(c("census", class(combined_data)))
  
  # return combined tibble
  combined_data
}
```

### Test Section

```{r}
# test combining two years
census_years_test <- census_query_years(years = c(2023, 2024))

# inspect results
# head() shows the first six rows
head(census_years_test)
# str() shows the structure and variable types
str(census_years_test)
# class() shows the classes attached to the object
class(census_years_test)
# unique() shows the distinct survey years included
unique(census_years_test$year)

```
The test shows that the 2023 and 2024 data were successfully combined into one tibble, with the year variable identifying the survey year for each row. The combined object also retains the "census" class so it can be used with the custom summary and plotting functions created next.

# Part 2: Writing Generic Functions

## Numeric and Categorical Summaries

Now that we have functions that return Census data for one or multiple years, we can create a custom `summary()` function for objects with the `"census"` class. Because we added `"census"` to the class of our returned tibbles, R can use a method called `summary.census()` whenever `summary()` is called on one of these objects.

Since each row does not necessarily represent one person, we also need to use `PWGTP` when calculating summaries. The function will calculate weighted means and standard deviations for numerical variables and weighted counts for categorical variables. Time variables will be excluded from the numerical summaries, and `year` will be treated as categorical.

```{r}
# custom summary method for objects with the "census" class
summary.census <- function(
  object,
  numerical_variables = NULL,
  categorical_variables = NULL
) {
  
  # identify time variables to exclude from numerical summaries
  time_variables <- c("JWAP", "JWDP")
  
  # if numerical variables are not specified, use all numeric variables
  if (is.null(numerical_variables)) {
    
    # sapply() checks each column to determine whether it is numeric
    numerical_variables <- names(object)[
      sapply(object, is.numeric)
    ]
    
    # remove PWGTP, year, and time variables from numerical summaries
    numerical_variables <- setdiff(
      numerical_variables,
      c("PWGTP", "year", time_variables)
    )
  }
  
  # if categorical variables are not specified, use factor and character variables
  if (is.null(categorical_variables)) {
    
    # identify columns that are stored as factors or character values
    categorical_variables <- names(object)[
      sapply(
        object,
        function(x) is.factor(x) || is.character(x)
      )
    ]
    
    # treat year as a categorical variable for summary purposes
    if ("year" %in% names(object)) {
      categorical_variables <- c(categorical_variables, "year")
    }
  }
  
  # create an empty list to store summary results
  summary_results <- list()
  
  # calculate weighted summaries for numerical variables
  for (variable in numerical_variables) {
    
    # pull numerical values and population weights
    numeric_values <- object[[variable]]
    weight_values <- object$PWGTP
    
    # is.na() identifies missing values
    # keep only rows where both the variable and weight are not missing
    valid_rows <- !is.na(numeric_values) & !is.na(weight_values)
    
    numeric_values <- numeric_values[valid_rows]
    weight_values <- weight_values[valid_rows]
    
    # calculate weighted mean using PWGTP
    weighted_mean <- sum(
      numeric_values * weight_values
    ) / sum(weight_values)
    
    # calculate weighted standard deviation using PWGTP
    weighted_sd <- sqrt(
      sum(numeric_values^2 * weight_values) / sum(weight_values) -
        weighted_mean^2
    )
    
    # save the mean and standard deviation for this variable
    summary_results[[variable]] <- c(
      mean = weighted_mean,
      sd = weighted_sd
    )
  }
  
  # calculate weighted counts for each categorical variables
  for (variable in categorical_variables) {
    
    # pull categorical values and population weights
    category_values <- object[[variable]]
    weight_values <- object$PWGTP
    
    # keep only rows without missing category values or weights
    valid_rows <- !is.na(category_values) & !is.na(weight_values)
    
    category_values <- category_values[valid_rows]
    weight_values <- weight_values[valid_rows]
    
    # tapply() splits the weights by category and applies sum() to each group
    weighted_counts <- tapply(
      weight_values,
      category_values,
      sum
    )
    
    # save weighted counts for this variable
    summary_results[[variable]] <- weighted_counts
  }
  
  # return the results as a named list
  summary_results
}
```

### Test Section 

```{r}
# test custom summary method using the default variables
summary(census_years_test)

# test summary method with selected variables
summary(
  census_years_test,
  numerical_variables = "AGEP",
  categorical_variables = "SEX"
)

```
The default summary calculates the weighted mean and standard deviation for AGEP and weighted counts for the categorical variables in the tibble, including year. For the combined 2023 and 2024 data, the weighted mean age is approximately 39.77 years. The second call demonstrates that the user can also specify which numerical and categorical variables to summarize, in this case returning only AGEP and SEX.


## Graphical Summaries

Next, we create a custom `plot()` function for Census objects. Because our returned tibbles have the `"census"` class, calling `plot()` on one of these objects will use our `plot.census()` method. The function will allow the user to specify one categorical variable and one numerical variable and will create a weighted boxplot using `PWGTP`.

```{r}
#custom plot method for objects with the "census" class
plot.census <- function(
  x,
  cat_var,
  num_var,
  ...
) {
  
  #ggplot() creates the plot using the census tibble x
  ggplot(
    x,
    aes(
      # get() allows us to use column names supplied as character strings
      # factor() makes sure the x-axis variable is treated as categorical
      x = factor(get(cat_var)),
      # get() identifies the selected numerical variable
      y = get(num_var),
      # PWGTP weights each observation by the population it represents
      weight = PWGTP
    )
  ) +
    # geom_boxplot() creates a boxplot for each category
    geom_boxplot() +
    labs( # labs() adds axis labels and a title
      x = cat_var,
      y = num_var,
      title = paste(num_var, "by", cat_var)
    )
}
```

### Test Section
```{r}
# test plot method using year as the categorical variable
# and age as the numerical variable
plot(
  census_years_test,
  cat_var = "year",
  num_var = "AGEP"
)

# test plot method using SEX as the categorical variable
plot(
  census_years_test,
  cat_var = "SEX",
  num_var = "AGEP"
)
```

The plot compares the weighted distribution of age between the 2023 and 2024 Census data. Using year as the categorical variable demonstrates that the year variable can be treated as a category even though it is stored numerically in the tibble. The second plot uses SEX instead of year, showing that the same plotting function can be used with different categorical inputs.

# Part 3: Investigation

Now that we have created and tested the API, summary, and plotting functions, we can use them to investigate a couple of questions from the PUMS data. We will look at changes from 2021 through 2024 and use `PWGTP` so that the summaries and plots account for the population represented by each record.

We will investigate two questions:

1. How has travel time to work changed from 2021 through 2024 in New York, and does it differ between males and females?
2. How have reported monthly household gas costs changed from 2021 through 2024 in Illinois?

## Travel Time to Work in New York

For the first investigation, we examine travel time to work in New York from 2021 through 2024 and compare the results for males and females. We use `JWMNP`, which measures travel time to work in minutes, along with `SEX` and `PWGTP`.
```{r}
# query New York travel time data for 2021 through 2024
commute_data <- census_query_years(
  years = c(2021, 2022, 2023, 2024),
  numerical_variables = c("JWMNP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = "36"
)

# inspect the returned data
head(commute_data)
nrow(commute_data)

# remove records where travel time to work is not applicable
commute_data_valid <- commute_data |>
  filter(JWMNP > 0)

# check the number of valid observations
nrow(commute_data_valid)

# export the data used for the analysis to Excel
writexl::write_xlsx(commute_data_valid, "data/commute_data_valid.xlsx")
```
The original query returned 820,516 observations. Because `JWMNP` contains a value of 0 for records where travel time to work is not applicable, we remove those observations before calculating the commute-time summaries. After filtering, 319,167 sampled records remain for the analysis.

Readers can inspect the data used for this investigation here: [Download the New York commute data](data/commute_data_valid.xlsx)

Next, we use our custom `summary()` method to summarize travel time and examine the weighted counts by sex and year.
```{r}
# summarize valid travel time observations
summary(
  commute_data_valid,
  numerical_variables = "JWMNP",
  categorical_variables = c("SEX", "year")
)
```
The weighted mean travel time across all four years is approximately 32.66 minutes, with a weighted standard deviation of approximately 25.74 minutes. The summary also provides weighted population counts by sex and year.

We first use our custom `plot()` method to compare the weighted distribution of travel time between males and females.
```{r}
# use our custom plot method to compare travel time by sex
plot(
  commute_data_valid,
  cat_var = "SEX",
  num_var = "JWMNP"
)
```
The boxplot shows substantial overlap in the travel-time distributions for males and females, although the typical travel time for males appears somewhat higher. Both groups also contain observations with considerably longer travel times.

The boxplot compares males and females across all four years combined. To examine how the difference changes over time, we calculate the weighted mean travel time separately for males and females within each year and then illustrate with a time series using `ggplot()`.

```{r}
# create an empty data frame to store the results
commute_results <- data.frame()

# calculate summaries for each year
for (current_year in c(2021, 2022, 2023, 2024)) {
  
  # calculate summaries separately for males and females
  for (current_sex in levels(commute_data_valid$SEX)) {
    
    # keep observations for one year and one sex
    commute_subset <- commute_data_valid |>
      filter(
        year == current_year,
        SEX == current_sex
      )
    
    # retain the census class
    class(commute_subset) <- unique(
      c("census", class(commute_subset))
    )
    
    # use our custom summary() method
    commute_summary <- summary(
      commute_subset,
      numerical_variables = "JWMNP",
      categorical_variables = "SEX"
    )
    
    # add the results to the results data frame
    commute_results <- bind_rows(
      commute_results,
      data.frame(
        year = current_year,
        SEX = current_sex,
        mean_travel_time =
          unname(commute_summary$JWMNP["mean"]),
        sd_travel_time =
          unname(commute_summary$JWMNP["sd"])
      )
    )
  }
}

# display the results
commute_results
```
The results show that males had a higher weighted mean travel time than females in each year from 2021 through 2024. In 2021, the weighted mean travel time was approximately 32.75 minutes for males and 29.93 minutes for females. By 2024, these averages had increased to approximately 34.40 minutes and 32.03 minutes, for males and females, respectively. Although travel times did not increase every year, both groups show an overall increase between 2021 and 2024.

Finally, we plot the yearly weighted means to make the change over time easier to compare.

```{r}
# plot weighted mean travel time by year and sex
ggplot(
  commute_results,
  aes(
    x = year,
    y = mean_travel_time,
    group = SEX,
    color = SEX
  )
) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Weighted Mean Travel Time (Minutes)",
    title = "Travel Time to Work in New York by Sex, 2021-2024"
  )
```
The line plot makes the pattern over time more apparent. Mean travel time increased substantially for both groups from 2021 to 2022, decreased slightly in 2023, and increased again in 2024. Males had a higher weighted mean travel time in every year, with the difference generally remaining between approximately two and three minutes.

## Monthly Household Gas Costs in Illinois

For the second investigation, we examine how reported monthly household gas costs changed in Illinois from 2021 through 2024. We use `GASP`, which measures monthly household gas cost in dollars, along with `PWGTP`.

```{r}
# query Illinois monthly gas cost data for 2021 through 2024
gas_data <- census_query_years(
  years = c(2021, 2022, 2023, 2024),
  numerical_variables = c("GASP", "PWGTP"),
  categorical_variables = "SEX",
  geography = "State",
  geography_value = "17"
)

# inspect the returned data
head(gas_data)

# inspect GASP values before calculating summaries
summary(gas_data$GASP)

# inspect the smallest values
sort(unique(gas_data$GASP))[1:15]

# count missing and zero values
sum(is.na(gas_data$GASP))
sum(gas_data$GASP == 0, na.rm = TRUE)

# check total number of observations
nrow(gas_data)

# export the Illinois gas-cost data used for the analysis
writexl::write_xlsx(
  gas_data,
  "data/gas_data_illinois.xlsx"
)
```
The Illinois query returned 511,124 observations across the four survey years. Unlike the travel-time variable in the first investigation, `GASP` contains no missing or zero values, so no additional filtering is needed before calculating the gas-cost summaries.

Readers can inspect the data used for this investigation here: [Download the Illinois household gas-cost data](data/gas_data_illinois.xlsx)


```{r}
# summarize monthly household gas cost and weighted counts by year
summary(
  gas_data,
  numerical_variables = "GASP",
  categorical_variables = "year"
)
```
Using `PWGTP`, the weighted mean reported monthly household gas cost across the four years is approximately $97.02, 
with a weighted standard deviation of approximately $163.50. The summary also provides the weighted population counts 
represented by the sample in each year.

```{r}
# use our custom plot method to compare gas costs across years
plot(
  gas_data,
  cat_var = "year",
  num_var = "GASP"
)
```
The boxplots show that reported monthly household gas costs are strongly right-skewed in each year, with a number of relatively high-cost observations. The center of the distribution appears higher in 2022 and 2023 than in 2021, while 2024 remains above the 2021 level. Because the distributions contain substantial outliers, the yearly weighted means will provide a clearer way to compare the overall change over time.


```{r}
# create an empty data frame to store yearly results
gas_results <- data.frame()

# calculate a gas-cost summary for each year
for (current_year in c(2021, 2022, 2023, 2024)) {
  
  # keep observations for one year
  gas_subset <- gas_data |>
    filter(year == current_year)
  
  # retain the census class
  class(gas_subset) <- unique(
    c("census", class(gas_subset))
  )
  
  # use our custom summary() method
  gas_summary <- summary(
    gas_subset,
    numerical_variables = "GASP",
    categorical_variables = "year"
  )
  
  # add the result to the results data frame
  gas_results <- bind_rows(
    gas_results,
    data.frame(
      year = current_year,
      mean_gas_cost =
        unname(gas_summary$GASP["mean"]),
      sd_gas_cost =
        unname(gas_summary$GASP["sd"])
    )
  )
}

# display results
gas_results
```
The yearly summaries show that the weighted mean reported monthly household gas cost increased from approximately $74.71 in 2021 to $113.67 in 2022. The mean then declined to approximately $105.38 in 2023 and $94.52 in 2024. Although reported monthly gas costs decreased after reaching their highest level in 2022, the 2024 weighted mean remained higher than the 2021 level.

The relatively large standard deviations are also consistent with the wide spread and high-cost observations seen in the boxplots.

```{r}
# plot weighted mean monthly gas cost by year
ggplot(
  gas_results,
  aes(
    x = year,
    y = mean_gas_cost
  )
) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Weighted Mean Monthly Gas Cost ($)",
    title = "Monthly Household Gas Costs in Illinois, 2021-2024"
  )
```
The line plot makes the change over time easier to see. Reported monthly household gas costs increased considerably between 2021 and 2022, followed by declines in 2023 and 2024. Despite those declines, the weighted mean in 2024 remained above its 2021 level.

**THE END**
