---
title: "ST558 Project 1"
author: "Alfonso Straffon"
format: html
editor: visual
---


# Introduction
The goal of this project is to write functions that will *manipulate* and *process* 
data from the Census Public Use Microdata Sample (PUMS) API, notably generic functions 
that will combine, summarize, and plot the queried data. The project will use PUMS 
data from 2021 through 2024.

# Initial Setup
Before starting the actual project we first need to: 1) create a github repository, 
2) set up github pages, 3) cloned the repo locally as an R project, 4) create a `.qmd` 
file, and 5) push said changes back to github in order to create a webpage. This 
sequence of steps largely followed what we did in Homework 1. Lastly, in the code 
chunk below, we loaded the needed R packages to be used throughout the project. 

Now we are ready to begin!
```{r}
#| message: false
#| warning: false

#needed packages throughout project
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
```

# Part 1: Data Processing

## Parsing Variable Values

To download the data we will use the `GET()` function from the `httr` package to 
query the URL for a year of interest, then we will parse the data with the `content()` 
function and store it as an R object (list), and lastly will select variables of 
interest. This same process will be repeated for the years 2021, 2022, 2023, and 
2024. 

```{r}

# Year 2021
# Get 2021 Census variable information
variable_return_2021 <- GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2021 <- content(variable_return_2021)

# Variables needed for 2021
variables_2021 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","ST", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2021 <- parsed_2021$variables[variables_2021]

# Pull the values for each variable
variable_values_2021 <- lapply(variable_info_2021, function(x) x$values)

# Year 2022
# Get 2022 Census variable information
variable_return_2022 <- GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2022 <- content(variable_return_2022)

# Variables needed for 2022
variables_2022 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","ST", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2022 <- parsed_2022$variables[variables_2022]

# Pull the values for each variable
variable_values_2022 <- lapply(variable_info_2022, function(x) x$values)

# Year 2023
# Get 2023 Census variable information
variable_return_2023 <- GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2023 <- content(variable_return_2023)

# Variables needed for 2023
variables_2023 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","STATE", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2023 <- parsed_2023$variables[variables_2023]

# Pull the values for each variable
variable_values_2023 <- lapply(variable_info_2023, function(x) x$values)

# Year 2024
# Get 2024 Census variable information
variable_return_2024 <- GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

# Parse JSON into an R object
parsed_2024 <- content(variable_return_2024)

# Variables needed for 2024
variables_2024 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP","PWGTP", "FER",
  "HHL", "SCH", "SCHL", "SEX","STATE", "REGION", "DIVISION")

# Keep only the variables needed for the project
variable_info_2024 <- parsed_2024$variables[variables_2024]

# Pull the values for each variable
variable_values_2024 <- lapply(variable_info_2024, function(x) x$values)

```

